Gravitational Lens Modeling in a Citizen Science Context

### 2014-09-16 text/ms.tex
more

\def\half{{\textstyle\frac12}}
% makes it easier to have consistent writing, and easy change if needed
\newcommand{\spl}{SpaghettiLens\xspace}
\newcommand{\sw}{Space Warps\xspace}

@@

The first two discoveries of gravitational lenses
\citep{1979Natur.279..381W,1980Natur.285..641W}, where a galaxy images
a single background quasar into two (four), brought in their wake the
first
work on lens modelling
\citep{1981ApJ...244..723Y,1981ApJ...244..736Y}.  For the second lens
(PG1115+080), mass models scored an early success with the prediction
that one of the lensed images seen would split further into two at
higher resolution.  That galaxies must sometimes cause multiple images
had long been expected \citep{1937ApJ....86..217Z}, and it had even
been argued that the phenomenon could help measure cosmological
parameters \citep{1964MNRAS.128..307R,1966MNRAS.132..101R}, but
apparently nobody was expecting that lenses would need detailed
modelling.  The first observations, however, immediately stimulated
models.  The reason for that lies in the image separation. Recall that
image separations are of order the angular Einstein radius
\begin{equation}
\sim \left(\frac{4GM}{c^2 D_L}\right)^{1/2}
\simeq 0.1'' \left(\frac{M}{M_\odot}\right)^{1/2}
             \left(\frac{D_L}{\rm pc}\right)^{-1/2}
\end{equation}
where $D_L$ is the distance to the lens, and $M$ its mass.  A lensing
galaxy with $M\sim10^{11}M_\odot$ at $10^9\rm\,pc$ would cause image
separations of $\sim1''$, which is comparable to the size of the
galaxy; typically the lensed images are seen through the galaxy halo.
Hence, the lensed images depend on the detailed mass distribution of
the lensing galaxy.  Galaxy lenses demand modelling.

@@

Imaging surveys now under way aim to increase the inventory of lenses
another ten or a hundred fold \citep[see
e.g.][]{Marshall2005,OguriMarshall2010}, with both automated and
visual search techniques proposed
\citep[e.g.][]{Marshall2009,More2012ApJ,Gavazzi2014}. For example, \sw
(Marshall et al, in prep, More et al, in prep) is a citizen science
project\footnote{\tt http://www.spacewarps.org} in which volunteers
are presented sky-survey images and invited to identify lens
candidates, by eye.  Simulated lenses are mixed in with the data,
both to help train volunteers on what to look for, and to provide
measures of the effectiveness of the search.  The motivation for \sw
is to enable volunteers, some of whom had previously serendipitously
identified lens candidates on earlier citizen-science surveys,
either to make discoveries missed in automatic searches by
software robots, or to perform the necessary inspection of an
automatically-generated sample, for quality control.  Robots can be
built to be good at detecting clean lensing system in uncrowded fields
with high signal-to-noise, but in more general test situations, robots
miss lenses (low completeness) or contaminate the results with
non-lenses (low purity) \citep{Marshall2009}.

@@

Consider a lens at some redshift $z_L$ and let $(x,y)$ be planar
coordinates at the lens, transverse to the line of sight.  Let
$\Sigma(x,y)$ be the mass per unit area, i.e., density projected along
the line of sight.  This surface density is often given in a
dimensionless form
\begin{equation} \label{eq:kappa}
\kappa(x,y) \propto \Sigma(x,y)
\end{equation}
called the convergence.  And let there be light, in the form of a more
distant source, at redshift $z_S$, behind point $(x_s,y_s)$ on the
lens.

>>>>

\def\half{{\textstyle\frac12}}
% makes it easier to have consistent writing, and easy change if needed
\newcommand{\spl}{SpaghettiLens\xspace}
\newcommand{\sw}{SpaceWarps\xspace}

@@

The first two discoveries of gravitational lenses
\citep{1979Natur.279..381W,1980Natur.285..641W}, where a galaxy images
a single background quasar into two (four), brought in their wake the
first work on lens modelling
\citep{1981ApJ...244..723Y,1981ApJ...244..736Y}.

%CM:
%The first work on lens modelling (Young et al. 1981a,b) was developed 
%after the discovery of the first two gravitational lenses (Walsh et al. 
% 1979; Weymann et al. 1980), where a massive galaxy causes a background 
% quasar to be presented as two or four images.

For the second lens
(PG1115+080), mass models scored an early success with the prediction
that one of the lensed images seen would split further into two at
higher resolution.  That galaxies must sometimes cause multiple images
had long been expected \citep{1937ApJ....86..217Z}, and it had even
been argued that the phenomenon could help measure cosmological
parameters \citep{1964MNRAS.128..307R,1966MNRAS.132..101R}, but
apparently nobody was expecting that lenses would need detailed
modelling.  The first observations, however, immediately stimulated
models.  The reason for that lies in the image separation. Recall that
image separations are of order the angular Einstein radius
\begin{equation}
\sim \left(\frac{4GM}{c^2 D_L}\right)^{1/2}
\simeq 0.1'' \left(\frac{M}{M_\odot}\right)^{1/2}
             \left(\frac{D_L}{\rm pc}\right)^{-1/2}
\end{equation}
where $D_L$ is the distance to the lens, and $M$ its mass.  A lensing
galaxy with $M\sim10^{11}M_\odot$ at $10^9\rm\,pc$ would cause image
separations of $\sim1''$, which is comparable to the size of the
galaxy; typically the lensed images are seen through the galaxy halo.
Hence, the lensed images depend on the detailed mass distribution of
the lensing galaxy.  Galaxy lenses demand modelling.

@@

Imaging surveys now under way aim to increase the inventory of lenses
another ten or a hundred fold \citep[see
e.g.][]{Marshall2005,OguriMarshall2010}, with both automated and
visual search techniques proposed
\citep[e.g.][]{Marshall2009,More2012ApJ,Gavazzi2014}. For example, \sw
(Marshall et al, in prep, More et al, in prep) is a citizen science
project\footnote{\tt http://www.spacewarps.org} in which volunteers
are presented sky-survey images and invited to identify lens
candidates, by eye.  Simulated lenses are mixed in with the data,
both to help train volunteers on what to look for, and to provide
measures of the effectiveness of the search.  The motivation for \sw
is to enable volunteers, some of whom had previously serendipitously
identified lens candidates on earlier citizen-science surveys,
either to make discoveries missed in automatic searches by
software robots, or to perform the necessary inspection of an
automatically-generated sample, for quality control.  Robots can be
built to be good at detecting lensing system in clean, uncrowded fields
with high signal-to-noise, but in more general test situations, robots
miss lenses (low completeness) or contaminate the results with
non-lenses (low purity) \citep{Marshall2009}.

@@

Consider a lens at some redshift $z_L$ and let $(x,y)$ be planar
coordinates at the lens, transverse to the line of sight.  Let
$\Sigma(x,y)$ be the mass per unit area, i.e., density projected along
the line of sight.  This surface density is often given in a
dimensionless form
\begin{equation} \label{eq:kappa}
\kappa(x,y) \propto \Sigma(x,y)
\end{equation}
called the convergence.  Let there be light, in the form of a more
distant source, at redshift $z_S$, behind point $(x_s,y_s)$ on the
lens.
